### 2023-11-09 main.tex
S5-shorter

 Using the conditional probability and the fact that the reversible measure \eqref{reversibleMeasure} is product over the sites of the chain of multinomial distributions we have

@@

Moreover, it holds that 
\text{Multinomial}(\nu,\rho_{1},\ldots,\rho_{N})+\text{Multinomial}(\nu,\rho_{1},\ldots,\rho_{N})=\text{Multinomial}(2\nu,\rho_{1},\ldots,\rho_{N})

@@

\paragraph{Proof of thermalized duality}

@@

We only report the final result. The details are written in Appendix \ref{appendix-RD}. For arbitrary site $x\in \{2,\ldots,L-1\}$, the action on the occupation variable $f(\bm{n})=n_{d}^{x}$ of particle of type $d\in \{1,\ldots,N-1\}$ reads

@@

	\mathcal{L}^{rd}n_{d}^{x}=\nu\sigma_{11}\Delta_{L}n_{d}^{x}+\nu\sigma_{12}\sum_{b=1\,:\,b\neq d}^{N-1}\Delta_{L}n_{c}^{x}+\Upsilon\sum_{b=1}^{N-1}(n_{b}^{x}-n_{d}^{x})\co
where $\Delta_{L}n_{x}^{d}= n_{x+1}^{d}+n_{x-1}^{d}-2n_{x}^{d}$. \\For $x=1$ and for all $d\in \{1,\ldots,N-1\}$ we have 

@@

	\mathcal{L}^{rd}n_{d}^{1}=\sigma_{11}\nu \left(\rho_{d}^{0}-2n_{d}^{1}+n_{d}^{2}\right)+\sigma_{12}\nu\sum_{b=1\,:\,b\neq d}^{N-1}\left(\rho_{b}^{0}-2n_{b}^{1}+n_{b}^{2}\right)+\Upsilon\sum_{b=1\,:\,b\neq d}^{N-1}\left(n_{b}^{1}-n_{d}^{1}\right)
On the occupation variable $n_{d}^{L}$ the action is similar, but with the upper case $1$ replaced by $L$.\\ In the spirit of \cite{casini2022uphill} we can define the average density occupation of species $d$ at site $x$ as
	\rho_{d}(x,t)=\mathbb{E}_{\mu}\left[n_{d}^{x}(t)\right]

@@

	\frac{d \rho_{d}(x,t)}{dt}=\mathbb{E}_{\mu}\left[\mathcal{L}^{rd}n_{d}^{x}(t)\right]

@@

\subsubsection{Reversible measure for the reaction-diffusion process} 
The process described by the generator \eqref{RDGenerator} is reversible with respect to the homogeneus product measure \begin{equation}\label{reversibleMeasureRD}
	\Lambda_{rev}=\bigtimes_{x=1}^{L}\Lambda_{rev}^{x}
\end{equation}
when for any $a\in\{1,\ldots,N\}$
\begin{equation}\label{reversibilityConditionRD}
	\alpha_{a}=\beta_{a}\qquad\text{and}\qquad \alpha_{a}=\alpha_{b}=\alpha \quad \forall a,b\in \{1,\ldots,N-1\}
\end{equation}
This measure has marginals $\Lambda_{rev}^{x}$ given by 
\begin{equation}
	\Lambda^{x}_{rev}\sim \text{Multinomial}\left(\twoj,p_{1},\ldots,p_{N}\right)
\end{equation}
where 
\begin{equation}
	p=p_{a}=\frac{\alpha}{\alpha_{N}+(N-1)\alpha}\quad \forall a\in \{1,\ldots,N-1\}\qquad \text{and}\qquad p_{N}=\frac{\alpha_{N}}{\alpha_{N}+(N-1)\alpha}
\end{equation}
Namely,
\begin{equation}
	\Lambda_{rev}^{x}(n^{x})=\frac{\nu!}{\prod_{a=1}^{N}n_{a}^{x}!}p^{\sum_{a=1}^{N-1}n_{a}^{x}}p^{n_{N}^{x}}
\end{equation}
This can be proved just by imposing the detailed balance conditions for the edge $x,x+1$ and for the boundaries. %For fixed site $x\in \{1,\ldots,L\}$ we only prove the detailed balance for  $\mathcal{L}_{x,x+1}^{c}$ with $c\in\{1,\ldots,N-2\}$ since for the generator $\mathcal{L}_{x,x+1}^{m}$ the computation is similar and has already been proved for $\mathcal{L}_{x,x+1}$ of the multi-species stirring. Consider the transition $\bm{n}\to \bm{n}-\delta_{a}^{x}+\delta_{g_{c}(b)}^{x}+\delta_{g_{c}(b)}^{x+1}-\delta_{b}^{x+1}$ for arbitrary $a,b\in\{1,\ldots,N-1\}$ and $c\in\{1,\ldots,N-2\}$; the detailed balance gives
\begin{comment}
\begin{equation}
	\begin{split}
		&n_{a}^{x}n_{b}^{x+1}\frac{\nu!}{n_{1}^{x}!\cdots n_{a}^{x}!\cdots n_{g_{c}(b)}^{x}!\ldots n_{N}^{x}!}\frac{\nu!}{n_{1}^{x+1}!\cdots n_{b}^{x+1}!\cdots n_{g_{c}(a)}^{x+1}!\ldots n_{N}^{x+1}!}\\&\rho_{1}^{n_{1}^{x}+n_{1}^{x+1}}\cdots \rho_{a}^{n_{a}^{x}}\cdots\rho_{g_{c}(b)}^{n_{g_{c}(b)}^{x}}\cdots\rho_{b}^{n_{b}^{x+1}}\cdots \rho_{g_{c}(a)}^{n_{g_{c}(a)}^{x+1}}\cdots \rho_{N}^{n_{N}^{x}+n_{N}^{x+1}}
		\\
		=&(n_{g_{c}(b)}^{x}+1)(n_{g_{c}(a)}^{x+1}+1)\frac{\nu!}{n_{1}^{x}!\cdots (n_{a}^{x}-1)!\cdots (n_{g_{c}(b)}^{x}+1)!\ldots n_{N}^{x}!}\frac{\nu!}{n_{1}^{x+1}!\cdots (n_{b}^{x+1}-1)!\cdots (n_{g_{c}(a)}^{x+1}+1)!\ldots n_{N}^{x+1}!}
		\\&
		\rho_{1}^{n_{1}^{x}+n_{1}^{x+1}}\ldots \rho_{a}^{n_{a}^{x}-1}\cdots\rho_{g_{c}(b)}^{n_{g_{c}(b)}^{x}+1}\cdots\rho_{b}^{n_{b}^{x+1}-1}\cdots \rho_{g_{c}(a)}^{n_{g_{c}(a)}^{x+1}+1}
	\end{split}
\end{equation}
that simplifies to

@@

@@

\paragraph{Action of $\mathcal{L}_{x,+1}^{rd}$ on the occupation variable}

>>>>

 Using the conditional probability and the fact that the reversible measure \eqref{reversibleMeasure} is the product over the sites of multinomial distributions we have

@@

%Moreover, it holds that 
%\text{Multinomial}(\nu,\rho_{1},\ldots,\rho_{N})+\text{Multinomial}(\nu,\rho_{1},\ldots,\rho_{N})=\text{Multinomial}(2\nu,\rho_{1},\ldots,\rho_{N})

@@

The proof duality result for the thermalized multi-species stirring process is standard. For the sake of completeness we report it in Appendix \ref{appendix-dualityThermalized}.

@@

We only report the final result. The details are written in Appendix \ref{appendix-RD}. For arbitrary site $x\in \{2,\ldots,L-1\}$, the action on the occupation variable $f(\bm{n})=n_{d}^{x}$ of particle of type $a\in \{1,\ldots,N-1\}$ reads

@@

	\mathcal{L}^{rd}n_{a}^{x}=\nu\sigma_{11}\Delta_{L}n_{a}^{x}+\nu\sigma_{12}\sum_{b=1\,:\,b\neq a}^{N-1}\Delta_{L}n_{b}^{x}+\Upsilon\sum_{b=1}^{N-1}(n_{b}^{x}-n_{a}^{x})\co
where $\Delta_{L}n_{a}^{x}= n_{a}^{x+1}+n_{x}^{x-1}-2n_{a}^{x}$. \\For $x=1$ and for all $a\in \{1,\ldots,N-1\}$ we have 

@@

	\mathcal{L}^{rd}n_{a}^{1}=\sigma_{11}\nu \left(\rho_{a}^{0}-2n_{a}^{1}+n_{a}^{2}\right)+\sigma_{12}\nu\sum_{b=1\,:\,b\neq a}^{N-1}\left(\rho_{b}^{0}-2n_{b}^{1}+n_{b}^{2}\right)+\Upsilon\sum_{b=1\,:\,b\neq d}^{N-1}\left(n_{b}^{1}-n_{a}^{1}\right)
On the occupation variable $n_{a}^{L}$ the action is similar, but with the upper case $1$ replaced by $L$.\\ In the spirit of \cite{casini2022uphill} we can define the average density occupation of species $d$ at site $x$ as
	\rho_{a}(x,t)=\mathbb{E}_{\mu}\left[n_{a}^{x}(t)\right]

@@

	\frac{d \rho_{a}(x,t)}{dt}=\mathbb{E}_{\mu}\left[\mathcal{L}^{rd}n_{a}^{x}(t)\right]

@@

\subsubsection{Reversible measure for the reaction-diffusion process (equilibrium)} 
The process described by the generator \eqref{RDGenerator} is reversible with respect to the homogeneus product measure \begin{equation}\label{reversibleMeasureRD}
	\Lambda_{rev}=\bigtimes_{x=1}^{L}\Lambda_{rev}^{x}
\end{equation}
when for any $a\in\{1,\ldots,N\}$
\begin{equation}\label{reversibilityConditionRD}
	\alpha_{a}=\beta_{a}\qquad\text{and}\qquad \alpha_{a}=\alpha_{b}=\alpha \quad \forall a,b\in \{1,\ldots,N-1\}
\end{equation}
This measure has marginals $\Lambda_{rev}^{x}$ given by 
\begin{equation}
	\Lambda^{x}_{rev}\sim \text{Multinomial}\left(\twoj,p_{1},\ldots,p_{N}\right)
\end{equation}
where 
\begin{equation}
	p=p_{a}=\frac{\alpha}{\alpha_{N}+(N-1)\alpha}\quad \forall a\in \{1,\ldots,N-1\}\qquad \text{and}\qquad p_{N}=\frac{\alpha_{N}}{\alpha_{N}+(N-1)\alpha}
\end{equation}
Namely,
\begin{equation}
	\Lambda_{rev}^{x}(n^{x})=\frac{\nu!}{\prod_{a=1}^{N}n_{a}^{x}!}p^{\sum_{a=1}^{N-1}n_{a}^{x}}p^{n_{N}^{x}}
\end{equation}
This can be proved just by imposing the detailed balance conditions for the edge $x,x+1$ and for the boundaries. %For fixed site $x\in \{1,\ldots,L\}$ we only prove the detailed balance for  $\mathcal{L}_{x,x+1}^{c}$ with $c\in\{1,\ldots,N-2\}$ since for the generator $\mathcal{L}_{x,x+1}^{m}$ the computation is similar and has already been proved for $\mathcal{L}_{x,x+1}$ of the multi-species stirring. Consider the transition $\bm{n}\to \bm{n}-\delta_{a}^{x}+\delta_{g_{c}(b)}^{x}+\delta_{g_{c}(b)}^{x+1}-\delta_{b}^{x+1}$ for arbitrary $a,b\in\{1,\ldots,N-1\}$ and $c\in\{1,\ldots,N-2\}$; the detailed balance gives
\begin{comment}
\begin{equation}
	\begin{split}
		&n_{a}^{x}n_{b}^{x+1}\frac{\nu!}{n_{1}^{x}!\cdots n_{a}^{x}!\cdots n_{g_{c}(b)}^{x}!\ldots n_{N}^{x}!}\frac{\nu!}{n_{1}^{x+1}!\cdots n_{b}^{x+1}!\cdots n_{g_{c}(a)}^{x+1}!\ldots n_{N}^{x+1}!}\\&\rho_{1}^{n_{1}^{x}+n_{1}^{x+1}}\cdots \rho_{a}^{n_{a}^{x}}\cdots\rho_{g_{c}(b)}^{n_{g_{c}(b)}^{x}}\cdots\rho_{b}^{n_{b}^{x+1}}\cdots \rho_{g_{c}(a)}^{n_{g_{c}(a)}^{x+1}}\cdots \rho_{N}^{n_{N}^{x}+n_{N}^{x+1}}
		\\
		=&(n_{g_{c}(b)}^{x}+1)(n_{g_{c}(a)}^{x+1}+1)\frac{\nu!}{n_{1}^{x}!\cdots (n_{a}^{x}-1)!\cdots (n_{g_{c}(b)}^{x}+1)!\ldots n_{N}^{x}!}\frac{\nu!}{n_{1}^{x+1}!\cdots (n_{b}^{x+1}-1)!\cdots (n_{g_{c}(a)}^{x+1}+1)!\ldots n_{N}^{x+1}!}
		\\&
		\rho_{1}^{n_{1}^{x}+n_{1}^{x+1}}\ldots \rho_{a}^{n_{a}^{x}-1}\cdots\rho_{g_{c}(b)}^{n_{g_{c}(b)}^{x}+1}\cdots\rho_{b}^{n_{b}^{x+1}-1}\cdots \rho_{g_{c}(a)}^{n_{g_{c}(a)}^{x+1}+1}
	\end{split}
\end{equation}
that simplifies to

@@

\section{Proof of duality of thermalized multi-species stirring}\label{appendix-dualityThermalized}

@@

\paragraph{Action of $\mathcal{L}_{x,x+1}^{rd}$ on the occupation variable}
